Median Maximum Pairwise Distance

### 2020-10-08 main.Rmd
updated structure to discuss

<!-- background of the problem -->
Take the example of the calendar display of electricity smart meter data used in @wang2020calendar for four households in Melbourne, Australia. The authors show how hour-of-the-day interact with weekday and weekends and then move on to use calendar display to show daily schedules. The calendar display has several components in it, which helps us look at energy consumption across hour-of-the-day, day-of-the-week, week-of-the-month, and month-of-the-year at once.
Some interaction of these cyclic granularities[@our-paper] could also be interpreted from this display. This is a great start to have an overview of the energy consumption. However, if one wants to understand the periodicities in energy behavior and how the periodicities interact in greater details, it is not easy to comprehend the interactions of some periodicities' from this display, due to the combination of linear and cyclic representation of time. For example, this display might not be the best to understand how hour-of-the-day varies across month-of-year varies across week-of-the-month. Further, it is not clear what all interactions of cyclic granularities should be read from this display as there could be many combinations that one can look at. Moreover, calendar effects are not restricted to conventional day-of-week or month-of-year deconstructions (@our-paper) and could include other cyclic granularities like hour-of-week or day-of-fortnight, which could potentially become useful depending on the context. Lastly, there might be specific interactions that are interesting and others that are not and that too will vary with different households. For example, area distribution quantiles are plotted for household 2 and 4 in Figure ref{fig:}a and b respectively. For the first household, the 75th and 90th percentile for Jan, Feb and July are very close, implying that energy usage for these months are generally on a much higher side due to the usage of air conditioners (in Jan and Feb) and heaters (in July).
The energy consumption for household 2 is also higher relative to its own consumption for Jan, Feb and March but the 75th and 90th percentile are apart implying that contrary to the first household, the second household resorts to air conditioners and heaters much less regularly than the first one. Moreover, the 75th percentile distribution is not bimodal across hours of the day for the first household in those months, but the distribution looks similar for all months for the second household. Difference in the energy consumption seem to be varying both across month-of-year (facets) and hour-of-day (x-axis). And thus, both the cyclic granularities would deem important while studying the periodicities in the first household. However, it seems like energy consumption across hours of the day are not that different across different months for the second household. Differences seem to be more prominent across month-of-year (facets) than hour-of-day (x-axis). Again, look at \ref{fig:} c and d, where energy consumption for these two households are plotted against (weekend/weekday, week-of-month). Here, for both households, the pattern of energy consumption vary  across different weeks of the month irrespective of the fact it is a weekday or weekend. In that respect, the harmony pair (month-of-year, hour-of-day) seems to be more informative than (weekend/weekday, week-of-month) for the first household. It could be immensely useful to make the transition from all possible ways to only ways that could potentially be informative given a household.

@@

# wknd_wday and week_month (important pair) id2's behavior across different hours of the day very different across months, but for id4 beahvior across different hours of the day is not likely a function of month.

@@

Here, we aim to build a new measure to follow through these two main objectives:

@@

<!-- To elaborate further, look at the examples in Figure \ref{}, where Figure \ref{}a represents the panel design with distribution of each x categories drawn from N(5, 10) distribution. It could be observed that the graph is not particularly interesting, as there is no significant change in distribution between x-axis levels or facets. Figure \ref{}b represents the same panel design with no difference in distribution of x-axis categories within a facet, but different distribution of x-axis categories for different facets. For example, if there are 4 facet levels and 2 x-axis levels, data is generated in the way as described in Table \ref{}. Figure \ref{}b exhibits an exact opposite situation where the x-axis within facets are different but not across facets. -->
<!-- Figure \ref{}d takes it futher by varying the distribution across both facet and x-axis categories. -->

@@

Suppose that $X_1$, $X_2$ , ... , $X_n$ are i.i.d. random variables with expected values $E(X_i) = \mu < \infty$ and variance $Var(X_i) = \sigma^2 < \infty$. Let 
$Y = max(X_1, X_2, \dots, X_n)$.

Let $F_X(x)$ be the common distribution of the variables $X_i$ and let $F_Y(y)$ be the corresponding distribution of $Y$. $F_Y(y)$ could be obtained from $F_X(x)$ simply by using:
$F_Y(y) = P[(X_1 \leq y)\cap(X_2 \leq y) \cap...\cap (X_n \leq y)] = {F_X(y)}^n$. For large $n$, the distribution of $Y$ approaches a standard shape, which does not depend on $F_X$. But what about the case when $n$ is not large enough? The distribution of maximum in that case will indeed depend on $n$ and the underlying distribution of $X$. If $F_X(x)$ is the CDF of $X$, then $F_Y(y) = {F_x(y)}^n$. Suppose $\Phi$ nd $\phi$ are the cdf and pdf of a standard normal distribution, then
$f_Y(y) = n{\Phi(y)}^{n-1}\phi(y)$, which depends on $n$. Hence, we are trying to normalise for $n$. Also, it depends on the underlying distribution of $X$, which we have assumed as normal in our case. As $n$ grows, we  can see the right tail growing, which implies that the probability that we will get a higher maximum is more. Now, for large $n$, we used EVT to normalise for $n$, that is, we brought them to the same scale without distorting the range of the distribution. But in our case, we will mostly have small $n$. It is important to ensure that they have the same mean and variation, for being able to compare the maximum value across $n$. We observe from the following graphs that our normalisation works after $n=6$, after which the difference in mean and standard deviation flattens out a lot.

Maximum pairwise distances are not robust to the number of levels and is higher for harmonies with higher levels. Thus these maximum pairwise distances need to be normalized for different harmonies in a way that eliminates the effect of different levels. The Fisher–Tippett–Gnedenko theorem in the field of Extreme Value Theory states that the maximum of a sample of iid random variables after proper re-normalization can converge in distribution to only one of Weibull, Gumbel or Freschet distribution, independent of the underlying data or process.

@@

### why normalize

@@

### Mean and standard deviation of the distribution of maximum

```{r}
library(ggridges)
library(viridis)
max_all_data <- create_nlevels(nlevels = seq(2, 30, 2),
                           nsim = 500,
                           sim_dist = distributional::dist_normal(5, 10), create_fun = max)

@@

## Null distribution of MMPD

@@

### Confidence interval of test statistic

## Simulation design

Behavior of the statistic - control simulation

 - To check if different distributions impact
(simulate with different distributions but same for all levels)

 - To check if x-levels and facets are normalized
(simulate with different distributions)
(simulate with different x and facet levels)

## Size and power

@@

```{r normalv21_power}

```{r normalv22_power}

### Power: Simulated same distribution for all combinations of categories for all harmony pairs.

```{r normalv23_power}

```{r normalv24_power}

*Conclusion*: The test rejects the null hypothesis if distributions are different.

### Scenario 2: Simulated different distributions for all combinations of categories for harmony pairs for few levels.

```{r diffnull_2by4, eval = FALSE}

*Conclusion*: The test select the harmony pair for which distribution of x-axis categories are significantly different

### Scenario 3: Simulated different distributions for all combinations of categories for all harmony pairs with many levels.

```{r diffnull_7by11, eval = FALSE}
```

*Conclusion*: The test indicates that both harmony pairs do not have significant variation.

### Scenario 4: Simulated different distributions for all combinations of categories for all harmony pairs with many levels - very different distribution across x-axis

```{r diffnull_7by11normal, eval = FALSE}
```

*Conclusion*: The test indicates that only the first harmony pair has significant variation.

### Scenario 5: Simulated different distributions for all combinations of categories for all harmony pairs with many levels - very different distribution across facets

```{r diffnull_7by11normal2, eval = FALSE}
```

@@

<!-- <!-- ```{r diffnull_3levels} --> -->

<!-- <!-- ``` --> -->

@@

Few harmony pairs are displayed in \autoref{fig:bothcust} to illustrate the siginificance of MMPD, threshold and the impact of different distribution plots and reverse mapping.
\noindent For each of \autoref{fig:bothcust} (b) and (c), $C_i$ is the circular granularity day-type (weekday/weekend) and $C_j$ is hour of the day. The geometry used for displaying the distribution is chosen as area-quantiles and violins in \autoref{fig:bothcust} (b and c respectively). \autoref{fig:bothcust} (a) displays reverse mapping of $C_i$ and $C_j$ with $C_i$ denoting hour of the day and $C_j$ denoting day-type with distribution geometrically displayed as boxplots.

>>>>

<!-- background of the problem -->
Take the example of the calendar display of electricity smart meter data used in @wang2020calendar for four households in Melbourne, Australia. The authors show how hour-of-the-day interact with weekday and weekends and then move on to use calendar display to show daily schedules. The calendar display has several components in it, which helps us look at energy consumption across hour-of-the-day, day-of-the-week, week-of-the-month, and month-of-the-year at once.
Some interaction of these cyclic granularities[@our-paper] could also be interpreted from this display. This is a great start to have an overview of the energy consumption. However, if one wants to understand the periodicities in energy behavior and how the periodicities interact in greater details, it is not easy to comprehend the interactions of some periodicities' from this display, due to the combination of linear and cyclic representation of time. For example, this display might not be the best to understand how hour-of-the-day varies and month-of-year varies across week-of-the-month. Further, it is not clear what all interactions of cyclic granularities should be read from this display as there could be many combinations that one can look at. Moreover, calendar effects are not restricted to conventional day-of-week or month-of-year deconstructions (@our-paper) and could include other cyclic granularities like hour-of-week or day-of-fortnight, which could potentially become useful depending on the context. Lastly, there might be specific interactions that are interesting and others that are not and that too will vary with different households. For example, area distribution quantiles are plotted for household 2 and 4 in Figure ref{fig:}a and b respectively. For the first household, the 75th and 90th percentile for Jan, Feb and July are very close, implying that energy usage for these months are generally on a much higher side due to the usage of air conditioners (in Jan and Feb) and heaters (in July).
The energy consumption for household 2 is also higher relative to its own consumption for Jan, Feb and March but the 75th and 90th percentile are apart implying that contrary to the first household, the second household resorts to air conditioners and heaters much less regularly than the first one. Moreover, the 75th percentile distribution is not bimodal across hours of the day for the first household in those months, but the distribution looks similar for all months for the second household. Difference in the energy consumption seem to be varying both across month-of-year (facets) and hour-of-day (x-axis). And thus, both the cyclic granularities would deem important while studying the periodicities in the first household. However, it seems like energy consumption across hours of the day are not that different across different months for the second household. Differences seem to be more prominent across month-of-year (facets) than hour-of-day (x-axis). Again, look at \ref{fig:} c and d, where energy consumption for these two households are plotted against (weekend/weekday, week-of-month). Here, for both households, the pattern of energy consumption vary  across different weeks of the month irrespective of the fact it is a weekday or weekend. In that respect, the harmony pair (month-of-year, hour-of-day) seems to be more informative than (weekend/weekday, week-of-month) for the first household. It could be immensely useful to make the transition from all possible ways to only ways that could potentially be informative given a household.

@@

# wknd_wday and week_month (important pair) id2's behavior across different hours of the day very different across months, but for id4 behavior across different hours of the day is not likely a function of month.

@@

In this paper, we aim to build a new measure to follow through these two main objectives:

@@

<!-- To elaborate further, look at the examples in Figure \ref{}, where Figure \ref{}a represents the panel design with distribution of each x categories drawn from N(5, 10) distribution. It could be observed that the graph is not particularly interesting, as there is no significant change in distribution between x-axis levels or facets. Figure \ref{}b represents the same panel design with no difference in distribution of x-axis categories within a facet, but different distribution of x-axis categories for different facets. For example, if there are 4 facet levels and 2 x-axis levels, data is generated in the way as described in Table \ref{}. Figure \ref{}b exhibits an exact opposite situation where the x-axis within facets are different but not across facets. -->
<!-- Figure \ref{}d takes it further by varying the distribution across both facet and x-axis categories. -->

@@

<!-- bullet points end here -->

<!-- Suppose that $X_1$, $X_2$ , ... , $X_n$ are i.i.d. random variables with expected values $E(X_i) = \mu < \infty$ and variance $Var(X_i) = \sigma^2 < \infty$. Let  -->
<!-- $Y = max(X_1, X_2, \dots, X_n)$. -->

<!-- Let $F_X(x)$ be the common distribution of the variables $X_i$ and let $F_Y(y)$ be the corresponding distribution of $Y$. $F_Y(y)$ could be obtained from $F_X(x)$ simply by using: -->
<!-- $F_Y(y) = P[(X_1 \leq y)\cap(X_2 \leq y) \cap...\cap (X_n \leq y)] = {F_X(y)}^n$. For large $n$, the distribution of $Y$ approaches a standard shape, which does not depend on $F_X$. But what about the case when $n$ is not large enough? The distribution of maximum in that case will indeed depend on $n$ and the underlying distribution of $X$. If $F_X(x)$ is the CDF of $X$, then $F_Y(y) = {F_x(y)}^n$. Suppose $\Phi$ nd $\phi$ are the cdf and pdf of a standard normal distribution, then -->
<!-- $f_Y(y) = n{\Phi(y)}^{n-1}\phi(y)$, which depends on $n$. Hence, we are trying to normalise for $n$. Also, it depends on the underlying distribution of $X$, which we have assumed as normal in our case. As $n$ grows, we  can see the right tail growing, which implies that the probability that we will get a higher maximum is more. Now, for large $n$, we used EVT to normalise for $n$, that is, we brought them to the same scale without distorting the range of the distribution. But in our case, we will mostly have small $n$. It is important to ensure that they have the same mean and variation, for being able to compare the maximum value across $n$. We observe from the following graphs that our normalisation works after $n=6$, after which the difference in mean and standard deviation flattens out a lot. -->

<!-- ### why normalize -->

Median and maximum distances are not robust to the number of levels and the metric would have higher values for harmonies with higher levels. Figure \ref{fig:mean-sd-max} shows how both mean and standard deviation of the distribution of maximum and median varies for different sample sizes ($n$). There is a rightward shift in the location and decrease in scale with increasing $n$. Thus these median maximum pairwise distances need to be normalized for different harmonies in a way that eliminates the effect of different levels, consequently enabling comparison across different harmonies. The Fisher–Tippett–Gnedenko theorem in the field of Extreme Value Theory states that the maximum of a sample of iid random variables after proper re-normalization can converge in distribution to only one of Weibull, Gumbel or Freschet distribution, independent of the underlying data or process.

@@

For large $n$, the distribution of $M$ approaches a standard shape. But what about the case when $n$ is not large enough? The distribution of maximum in that case will indeed depend on $n$ and the underlying distribution of $D$. For large $n$, we could use EVT and CLT to normalise for $n$, however, most of the cyclic granularities encountered in practice have low $n$. Hence, normalisation could be done by bootstrapping procedure to ensure that the distribution of maximum/median have the same mean and variation for varying $n$.

@@

```{r mean-sd-max, fig.cap = "Display shows mean and standard deviation of the distribution of maximum for sample sizes ranging from 2 to 30. There is a right shift in the distribution of maximum and median with increasing sample sizes."}

library(ggridges)
library(viridis)
max_all_data <- create_nlevels(nlevels = seq(2, 30, 2),
                           nsim = 500,
                           sim_dist = distributional::dist_normal(5, 10), create_fun = max)

@@

## Design and analysis of simulation experiments

Simulation experiments with different designs could help monitor and understand the behavior of the test statistic.

### Impact of different levels from the same distribution of the measured variable (for understanding size)

A 2 way contingency table with number of facet levels, x-axis levels should be constructed with values as the p-values.

Table
Conclusions
(in Null distribution)

### Impact of different distribution of the measured variable keeping the same distribution for each
panel (for understanding size)

To check if different distributions impact, simulate data from different distributions. A 3 way contingency table with number of facet levels, x-axis levels and distribution type should be constructed with each cell value as the p-value.

Table
Conclusions
(in Null distribution)

### Impact of simulating different distributions across categories for understanding power

#### Varying distribution across facet
#### Varying distribution across x-axis
#### Varying distribution across both facets and x-axis
#### Repeat all with varying facet and x-axis levels

Table
Conclusions
(in power)

## Null distribution

@@

<!-- ### Confidence interval of test statistic -->

## Power and confidence interval

Failure to reject the null hypothesis when there is in fact a significant effect.

@@

<!-- # ```{r normalv21_power} -->
<!-- #  -->
<!-- # ``` -->
<!-- #  -->
<!-- #  -->
<!-- # ```{r normalv22_power} -->
<!-- #  -->
<!-- # ``` -->

### Varying distribution across facet
### Varying distribution across x-axis
### Varying distribution across both facets and x-axis
### Repeat all with varying facet and x-axis levels

<!-- ### Power: Simulated same distribution for all combinations of categories for all harmony pairs. -->

<!-- # ```{r normalv23_power} -->
<!-- #  -->
<!-- # ``` -->
<!-- #  -->
<!-- #  -->
<!-- # ```{r normalv24_power} -->
<!-- #  -->
<!-- # ``` -->

*Conclusion*: The test should reject the null hypothesis if distributions are different.

<!-- ### Scenario 2: Simulated different distributions for all combinations of categories for harmony pairs for few levels. -->

<!-- ```{r diffnull_2by4, eval = FALSE} -->

<!-- *Conclusion*: The test select the harmony pair for which distribution of x-axis categories are significantly different -->

<!-- ### Scenario 3: Simulated different distributions for all combinations of categories for all harmony pairs with many levels. -->

<!-- ```{r diffnull_7by11, eval = FALSE} -->
<!-- ``` -->

<!-- *Conclusion*: The test indicates that both harmony pairs do not have significant variation. -->

<!-- ### Scenario 4: Simulated different distributions for all combinations of categories for all harmony pairs with many levels - very different distribution across x-axis -->

<!-- ```{r diffnull_7by11normal, eval = FALSE} -->
<!-- ``` -->

<!-- *Conclusion*: The test indicates that only the first harmony pair has significant variation. -->

<!-- ### Scenario 5: Simulated different distributions for all combinations of categories for all harmony pairs with many levels - very different distribution across facets -->

<!-- ```{r diffnull_7by11normal2, eval = FALSE} -->
<!-- ``` -->

@@

<!-- <!-- ```{r diffnull_3levels} -->

<!-- <!-- ``` -->

@@

Few harmony pairs are displayed in \autoref{fig:bothcust} to illustrate the significance of MMPD, threshold and the impact of different distribution plots and reverse mapping.
\noindent For each of \autoref{fig:bothcust} (b) and (c), $C_i$ is the circular granularity day-type (weekday/weekend) and $C_j$ is hour of the day. The geometry used for displaying the distribution is chosen as area-quantiles and violins in \autoref{fig:bothcust} (b and c respectively). \autoref{fig:bothcust} (a) displays reverse mapping of $C_i$ and $C_j$ with $C_i$ denoting hour of the day and $C_j$ denoting day-type with distribution geometrically displayed as boxplots.
